Co-citations in context

### 2019-05-17 cocit_r.tex
update

\section{Introduction}
Citation and network analysis of  scientific literature reveals information on semantic relationships between publications, collaboration between scientists, and the practice of citation itself~\citep{garfield_citation_1955,de_solla_price_networks_1965,newman_structure_2001,Shi:2010:CHI:1816123.1816131,patience_pmid28560354}. Co-citation, the frequency with which two documents are cited together in other documents~\citep{small_co-citation_1973,marshakova-shaikevich_co-citation_1973}, provides additional insights, including the identification of semantically related documents and fields. Co-citations, bibliographic coupling and direct citation have been compared for XX.

In an intriguing study, co-citation analysis of 17.9 million articles and their cited references from the Web of Science (WoS) was used to document novel combinations of prior research and to characterize highly cited papers~\citep{uzzi_atypical_2013}. In this study, pairwise combinations of the references in each article were generated to create a multi-set of journal-pairs. Observed frequencies of these journal pairs across the dataset were computed and normalized to expected values generated by Monte Carlo simulations from a random graph model. Normalized, journal pair frequencies were termed \emph{z-scores} (Materials and Methods). Thus, every article was associated with multiple z-scores. For each article, positional statistics of these z-scores were calculated to describe conventionality (high conventionality (HC) if the median z-score for an article was greater than the median of median z-scores of all articles and low conventionality for the inverse (LC).) Similarly with novelty (high novelty (HN)- if the tenth percentile of z-scores for an article was less than zero and low novelty (LN) for the inverse). Accordingly, each article was labeled with respect to conventionality and novelty, e.g, HCLN, with all four combinations being possible. The finding of considerable interest was that HCHN articles were twice as likely to be found in highly cited papers (defined as the top 10\% of all papers published in a given year considering the count of citations accumulated in the first 8 years since publication).

Key to the Uzzi study, however, is the random graph approach, where every reference is randomly replaced by another article published in the same year as the reference to generate a null model for comparison. However, the approach inadequately accounts for observed citation behavior~\citep{wallace_lariviere_gingras_2012,moed_measuring_2010,klavans_research_2017} since (i) it treats potential substitutions as equi-probable, irrespective of disciplinary origin (ii) does not consider the frequency with which references are cited in the dataset. In illustration, a reference from musicology can replace one in quantum physics during the simulation process. Similarly, a reference cited over 100 times in a given year is selected with the same probability as a reference cited only once. Accordingly, the expected value calculations used in normalizing observed journal-pair-frequencies can be reasonably questioned on grounds of model mis-specification.

A follow-up study using a smaller dataset of 12 million articles, also suggests that Uzzi and colleagues do not fully account for disciplinary and journal effects~\citep{boyack_vs_uzzi_2014}. While the two studies are conceptually similar, the authors of the latter study used Scopus data and the K50 measurement, a derivative of the $\chi^2$ method, rather than Monte Carlo simulations for normalizing observed to expected journal-pair frequencies. Boyack and Klavans note that `only 64.4\%  of  243  WoS  subject  categories' in the Uzzi study met the criterion of having the highest probability of hit papers in the HCHN category. Further, they observed that journals vary widely in terms of size and influence and that 20 journals accounted for nearly 15\% of co-citations in their measurements. Lastly, that three multidisciplinary journals accounted for 9.4\% of all atypical combinations, suggesting strong effects from both disciplines and journals.

Despite different methods used to generate expected values, both Uzzi and Boyack measured observed and expected journal-pair frequencies across disciplines and then examined subsets of publications for disciplinary effects. Since disciplinary origin was not considered when generating expected values in their studies, we chose instead to analyze semantically related sets of documents (disciplinary networks) instead of all publications in a bibliographic database. We used keyword searches of the scientific literature to create three citation networks themed around broad search terms. For each of these, we calculated expected values using a refined random graph model that accounted for existing citation frequencies in these networks and an efficient simulation algorithm that randomly permuted references grouped by year of publication. Thus, while Uzzi et al. averaged the results of 10 simulations for each year in their dataset, we averaged 1,000 simulations to ensure greater coverage of rarely observed journal pairs We hypothesized that our approach would  reduce model misspecification and better simulate citation practice, in turn leading to different and more relevant conclusions.  Our study on these semantically-themed citation networks reveals significantly different patterns of conventionality and novelty between  citation networks and disciplines that challenges the conclusion of universality in HCHN articles have a greater probability of being represented in highly cited publications.

\section{Materials and Methods}
\section{Results}
\section{Discussion}

>>>>

\section{Introduction}
Citation and network analysis of  scientific literature reveals information on semantic relationships between publications, collaboration between scientists, and the practice of citation itself~\citep{garfield_citation_1955,de_solla_price_networks_1965,newman_structure_2001,Shi:2010:CHI:1816123.1816131,patience_pmid28560354}. Co-citation, the frequency with which two documents are cited together in other documents provides additional insights, including the identification of semantically related documents, fields, idea, and specialization in science~\citep{small_co-citation_1973,marshakova-shaikevich_co-citation_1973}.

In an intriguing study, co-citation analysis of 17.9 million articles and their cited references from the Web of Science (WoS) was used to document novel combinations of prior research and to characterize highly cited papers~\citep{uzzi_atypical_2013}. In this study, pairwise combinations of the references in each article were generated to create a multiset of journal-pairs. Observed frequencies of these journal pairs across the dataset were computed and normalized to expected values generated by Monte Carlo simulations that shuffled cited references in the context of a random graph model. The resultant normalized, journal pair frequencies were termed \emph{z-scores} (Materials and Methods). Thus, every article was associated with multiple z-scores. For each article, positional statistics of these z-scores were calculated to describe conventionality;(high conventionality (HC) if the median z-score for an article was greater than the median of median z-scores of all articles and low conventionality for the inverse (LC). Similarly with novelty;  high novelty (HN) if the tenth percentile of z-scores for an article was less than zero and low novelty (LN) for the inverse. Accordingly, each article was labeled with respect to conventionality and novelty, e.g, HCLN, with all four combinations being possible. The finding of considerable interest was that HCHN articles were twice as likely to be found in highly cited papers~\citep{uzzi_atypical_2013} suggesting that novel ideas flavoring a base of conventional thought were a recipe for impact. The authors examined subsets of publications grouped by disciplined and asserted universality of this feature.

Key to the Uzzi study, however, is the random graph approach and the mechanism of citation shuffling employed. Random substitutions are made under conditions that preserve the number of publications in each year of the dataset as well as the number of references cited by each of these publications. However, the approach inadequately accounts for observed citation behavior~\citep{wallace_lariviere_gingras_2012,moed_measuring_2010,klavans_research_2017,garfield_1979} since (i) random substitutions are selected with equal probability irrespective of disciplinary origin (ii) replacements are drawn from the set of eligible references rather than the multi-set that accounts for the frequency with which a reference is cited. In illustration, a musicology reference from can replace a quantum physics reference during the simulation process. Similarly, a reference cited over 100 times in a given year is selected with the same probability as a reference cited only once. Accordingly, the expected value calculations generated by simulations and used in normalizing observed journal-pair-frequencies can be reasonably questioned on grounds of model mis-specification.

A follow-up study using a smaller dataset of 12 million articles, also suggests that Uzzi and colleagues do not fully account for disciplinary and journal effects~\citep{boyack_vs_uzzi_2014}. While the two studies are conceptually similar, the authors of the latter study used Scopus data and the K50 measurement, a derivative of the $\chi^2$ method, rather than Monte Carlo simulations for normalizing observed to expected journal-pair frequencies. Boyack and Klavans note that ``only 64.4\%  of  243  WoS  subject  categories'' in the Uzzi study met the criterion of having the highest probability of hit papers in the HCHN category. Further, they observed that journals vary widely in terms of size and influence and that 20 journals accounted for nearly 15\% of co-citations in their measurements. Lastly, that three multidisciplinary journals accounted for 9.4\% of all atypical combinations, suggesting strong effects from both disciplines and journals.

Despite different methods used to generate expected values, expected journal-pair frequencies were generated without disciplinary constraints in both these prior studies~\citep{uzzi_atypical_2013,boyack_vs_uzzi_2014}, after which subsets of publications were examined for disciplinary effects. To avoid the problem of discipline-insensitive substitutions, we chose instead to first construct semantically related sets of documents (disciplinary networks) and then measure observed and expected frequencies within these networks. Accordingly, we used keyword searches of the scientific literature to create three citation networks themed around broad search terms. For each of these, we calculated expected values using an efficient simulation algorithm that randomly permuted references grouped by year of publication in a random graph model that accounted for existing citation frequencies in these networks. We hypothesized that our approach would  reduce model misspecification and better simulate citation practice, in turn leading to different and more relevant conclusions. Additionally,while Uzzi et al. averaged the results of 10 simulations for each year in their dataset, we averaged the results of 1,000 simulations per dataset to ensure greater coverage of rarely observed journal pairs (Materials and Methods). Our study on these semantically-themed citation networks reveals significantly different patterns of conventionality and novelty between  citation networks and disciplines that challenges the conclusion of universality in HCHN articles have a greater probability of being represented in highly cited publications. Instead, we conclude that conventional thought as commonly defined in our study, Uzzi, and Boyack is more likely to drive higher citations.. \emph{these are just filler conclusions that must be carefully rewritten and added to}.

\section{Materials and Methods}

\subsection{Bibliographic data} We have previously developed ERNIE, infrastructure to parse the Web of Science (WoS) Core Collection into a relational database that supports analytical studies~\cite{Keserci371955}. WoS data stored in ERNIE spans the period 1900-2019 and consists of over 72 million publications. For this study, we generated an analytical dataset from years 1985 to 2005. The total number of publications in this dataset was just over 25 million publications (25,134,073).  For each of these years, we further restricted analysis to those of type Article. Since WoS data also contains incomplete references or references that point at other indexes,  we also considered only those references for which there were complete records (Table 1). Given consistent trends in the data, we analyzed flanking years and the mid-point (1985, 1995, and 2005). For example, WoS data for year 2005 contained 1,753,174 publications, which after restricting to type Article and considering only those references described above resulted in 916,573 publications, 6,095,594 unique references (set of references), and 17,167,347 total references (multiset of references).

% latex table generated in R 3.5.3 by xtable 1.8-4 package
% Fri May 17 14:28:31 2019
\begin{table}[ht]
\caption{Summary of WoS Analytical Dataset. UP: unique publications, UR: unique references, TR: total references. The number of publications, unique references, total references and the ratio of total references to unique references increases monotonically with each year indicating that both the number of documents and citation activity increase over time. Data for reference years is flanked by horizontal lines and shown in boldface. Only publications of type Article and references with complete WoS records are included in these counts.}
\centering
\begin{tabular}{|rrrrr|}
  \hline
Year & UP & UR & TR & TR/UR \\ 
  \hline
\textbf{1985} & \textbf{418495} & \textbf{2281297} & \textbf{5615496} & \textbf{2.46} \\ 
\hline
1986 & 402309 & 2316451 & 5708796 & 2.46 \\ 
1987 & 412936 & 2427347 & 5998513 & 2.47 \\ 
1988 & 426001 & 2545647 & 6354917 & 2.50 \\ 
1989 & 443144 & 2673092 & 6749319 & 2.52 \\ 
1990 & 458768 & 2827517 & 7209413 & 2.55 \\ 
1991 & 477712 & 2977784 & 7729776 & 2.60 \\ 
1992 & 492181 & 3134109 & 8188940 & 2.61 \\ 
1993 & 504488 & 3278102 & 8676583 & 2.65 \\ 
1994 & 523660 & 3458072 & 9255748 & 2.68 \\ 
 \hline
\textbf{1995} & \textbf{559685} & \textbf{3692575} & \textbf{9897946} & \textbf{2.68} \\ 
\hline
1996 & 663110 & 4144581 & 11641286 & 2.81 \\ 
1997 & 677077 & 4340733 & 12135104 & 2.80 \\ 
1998 & 693531 & 4573584 & 12728629 & 2.78 \\ 
1999 & 709827 & 4784024 & 13280828 & 2.78 \\ 
 2000 & 721926 & 5008842 & 13810746 & 2.76 \\ 
 2001 & 727816 & 5203078 & 14261189 & 2.74 \\ 
 2002 & 747287 & 5464045 & 15001390 & 2.75 \\ 
 2003 & 786284 & 5773756 & 16024652 & 2.78 \\ 
 2004 & 826834 & 6095594 & 17167347 & 2.82 \\ 
 \hline
\textbf{2005} & \textbf{916573} & \textbf{6629595} & \textbf{19066249} & \textbf{2.88} \\ 
 \hline
\end{tabular}
\end{table}

\subsection{Disciplinary datasets}

\section{Results}

(i) disciplinary networks are different
(ii) background makes a difference
(iii) 
(iv)

\section{Discussion}
